---
title: Vitamin D and Covid-19
author: Gordon Shotwell
date: '2020-05-03'
slug: vitamin-d-and-covid-19
categories: []
tags: []
lastmod: '2020-05-03T11:15:41-03:00'
layout: post
type: post
highlight: no
---

The New York Times had a [recent piece](https://www.nytimes.com/2020/05/03/world/asia/coronavirus-spread-where-why.html?referringSource=articleShare) about the puzzling ways that Covid-19 has spread across the world. 
In it they articulate a number of possible theories about why Covid-19 has effected some countries more than others, but I think they left out one potential explanation: Vitamin D status. 
```{r, include = FALSE}
library(dplyr)
df <- readr::read_csv("covid_04-28-2020.csv") %>% 
  dplyr::filter(Country_Region %in% c("Canada", "Australia"))

country_summary <- df %>% 
  group_by(Country_Region) %>% 
  summarize(cases = sum(Confirmed),
            recovered = sum(Recovered),
            active = sum(Active), 
            deaths = sum(Deaths)) %>% 
  mutate(population = c(25000000,
                        37000000)) %>% 
  mutate(deaths_per_million = round((deaths / population) * 1e6, 1))
  
```

# The story so far

So far the story we've been told about Covid-19 is that it is an extremely dangerous virus that spreads expoentially. 
The only way to contain this virus is to engage in fairly radical national shutdowns to flatten the curve and prevent transmission. 
Countries that take immediate, strong action, will have easier disease courses, while countries that don't will have overwhelemed health care systems and much higher death rates. 
I have been, and continue to be, a supporter of this story, but I have to admit that there are quite a few data points which really don't fit this narrative.

## A tale of two countries

Canada and Australia have had pretty similar Covid timelines. 
They both had their first case at the end of Janury, and hit a hundred cases around mid March. 
They are similarly sized countries with similar demographics, and both have robust testing infrastructure.
Despite this, Australia has 3.6 deaths per million people, while Canada's rate is twenty times as high. 

```{r, echo = FALSE}
country_summary %>% 
  select(country = Country_Region, cases, deaths, population, deaths_per_million) %>% 
  knitr::kable(format.argsgs = list(big_mark = ","))
```
The story that you're likely to hear from the Australian government is that their governmental response was timely and effective, and because of this they were able to prevent widespread infection. 
However, when you look at the timelines of policy changes between the two countries they're almost identical. 
They were both adjacent to large outbreaks.
Canada shares a border with the United States, while Australia had direct, unscreened flights from Wuhan until late January.
They both enacted screening and quarantine policies which were unable to wholy prevent disease importataion or community spread, 
and they both enacted similar social distancing regulations around the same time. 
For example Australia cancelled University classes in the third week of March, while most Canadian provinces closed schools in mid march.

Despite all this, the countries ended up with extreme differences in death rates. 
For example I live in Nova Scotia, a small province of 900,000 people, which has almost as many deaths (31) as New South Wales (39). 
NSW has eight times as many people as Nova Scotia, relatively higher density, and an outbreak which started two months earlier. 
So what the hell is going on?

There are a few good theories for the difference including:

- Heat preventing viral transmission 
- Australia's Aged Care system being better than Canada's
- Luck

One that hasn't gotten enough attention though is vitamin D status.
Vitamin D levels are lower in Canada than Australia in February and March because your skin can only manufacture the vitamin with relatively intense UV rays, and the sun is too weak to do that before around May in most Canadian provinces.

There's some preliminary (preprint) research that vitamin D is involved in Covid-19 mortality:

- A [restrospective study](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3571484) in the Phillipines found a significant association between vitamin D status and severe Covid infections.
- An [Indonesian](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3585561) study found that after controlling for age, gender, and comorbidity vitamin D deficiency as associated with a twelve times higher risk of death
- Covid-19 appears to [vary by latitude](https://www.ncbi.nlm.nih.gov/pubmed/32311755)
- 

Additionally there are some other factors that make vitamin D a plausible candidate:

- Vitamin D helps regulate the [renin angiotensin system](https://youtu.be/Nfp4X4O87DQ?t=1499) which includes the ACE2 receptor that Covid-19 binds to.
- Vitamin D deficiency is common among populations at risk for Covid-19 mortality, including the elderly, obese, health care workers, and those in institutional settings.
- Vitamin D reduces the risk of [other](https://www.bmj.com/content/356/bmj.i6583) acute respiratory infections. 

Lastly vitamin D would explain some other puzzling phenomena:

- Why hasn't the disease been worse in places like India or Africa with relative constrained public health resources?
- Why are people of colour at such a [high risk](https://www.theglobeandmail.com/world/article-racial-divide-of-covid-19-patients-in-us-grows-even-starker-as-new/) for severe Covid in northern latitudes, but not southern ones? Maybe this is racism, but don't they have racism in Australia and South Africa?
- Why is Lousiana doing better than New York? 
- With all the old people in Florida, why isn't Florida doing worse?

Now I'm not saying that vitamin D is definitely the cause of differential Covid-19 mortality,
it's quite possible, and even likely that there's some bundle of explanations for all of the above. But it would be very exciting it it were involved because it's an abundant, well understood therapy. 

## What I'm paying attention to?

If the vitamin D thesis is right, we should see a few things: 

1) Positive results from some of the controlled [clinical trials](https://clinicaltrials.gov/ct2/show/NCT04334005)
2) Southern US states with bad public policy responses like Florida and Georgia should see lower than expected death rates
3) By July, we should see Covid-19 outbreaks with high mortality rates in places like Argentina, Tazmania, and New Zealand




















