---
title: "Vitamin D and Covid: We don't need to wait for more data."
author: "Gordon Shotwell"
date: "2020-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



I recently posted this graph on Twitter which suggests that Covid-19 mortality is related to latitude:

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
df <- read_csv("april_28_deaths_latitude.csv")

df <- df %>% 
  mutate(Latitude = as.numeric(Latitude),
         deaths_per_million = as.numeric(deaths_per_million)) %>% 
  mutate(lat_group = dplyr::case_when(
    Latitude > 25 ~ "> 25 degrees",
    TRUE ~ " < 25 degrees"
  ))

plot_df <- df %>% 
  filter(deaths_per_million < 750)


plot_df %>% 
  ggplot(aes(x = lat_group, y = deaths_per_million, colour = Latitude)) +
  scale_color_viridis_c() +
  theme_minimal() +
  geom_jitter(alpha = 0.7, width = 0.2) +
  labs(title = "Covid-19 deaths per million",
         x = "Latitude",
         y = "Deaths per Million")
```

There's a particular type of person on the internet who sees a graph like this and reaches deep into their datascience bootcamp memories to exclaim "correlation doesn't imply causation!" or "there's no randomized clinical trial!" or even "there are differences in testing strategy!"

These responses are stupid not because they're wrong but because they ignore how decision making works. 
It would be great if we had several high quality clinical trials for every decision, 
but clinical trials are difficult, expensive, and take a long time to complete,
so we usually have to make medical decisions without clear clinical direction. 
Even in cases where there _is_ a lot of clinical data, that data may not map perfectly onto the actual decision. 
A drug may have been approved based on a large, multi-center randomized trial, but maybe you have to treat a pregnant diabetic, and the trial didn't include any of them.
As a result we usually have to make decisions with vague, provisional information. 

This is especially true in the case of Covid-19.
The disease has only existed for 9 months, so obviously we're not going to have a RCT for most medical interventions.
It would be a kind of malpractice to wait to treat patients, or engage in public health interventions, until we have clear guidance from a large, multi-center clinical trial.
We have to make these medical decisions today with today's knowledge, even if that knowledge is provisional. 

The way to make these decisions is by combining _prior knowledge_ with _risk assessment_. 
For example, we have a lot of prior knowledge about treating acutre respiratory distress with ventilators. 
Since Covid-19 looks like a respiratory disease, doctors have a prior expectation that mechanical ventilation will help Covid they enter respiratory distress.  
Note that we do this without a clinical trial, it's entirely possible that Covid-19 patients are [not helped](https://time.com/5820556/ventilators-covid-19/) by mechanical ventilation or that it's harmful, but if we wait to put people on ventilators until we have a Covid-19 ventilation study patients might die.
Similarly, public health interventions were put in place without perfect certainty about how Covid is transmitted. 
Public health officials didn't wait for a large clinical trial on social distancing. 
Instead they relied on their piror beliefs about how viruses like Covid are transmitted, 
and decided that the risks of letting the virus run are probably higher than putting in overly stringent public health interventions.

The second consideration is how the risk of action compares with that of inaction.
Given that don't know much about this virus, what's the downside risk if you're totally wrong about the intervention.
For example we don't really have good evidence that widespead mask usage prevents Covid-19, but since masks are definitely not harmful, and there's a plausible way they could reduce transmission we might decide to promote them. 
Similarly, we don't know that closing playgrounds reduces transmission, but closing them isn't going to hurt anyone so it's probably a good bet.

# Prior belief about Vitamin D and Covid

Our prior about Vitamin D and Covid should be based on the same logic as mechanical ventilation: 
Covid looks like a respiratory virus, so until we learn otherwise we should treat and prevent it like a respiratory virus. 
A [meta-analysis](https://www.sciencedirect.com/science/article/pii/S1871402120301156) found two important facts about vitamin D supplemntation and other respiratory conditions:

- Overall vitamin D supplementation reduced the incidence of respiratory conditions by 20%
- For the very deficient, supplementation reduced the incidence by around 70%

These results aren't that impressive for most respiratory viruses which have a low rate of complication. 
I might not get that excited about reducing my risk of dying from a cold from 0.05% to 0.04%, but those are big numbers for Covid.
If we could reduce the rate of Covid transmission by 20-40% that would both save a lot of lives, but also allow us to relax certain types of social distancing. 
For example if the virus has an R0 of 2.2 without supplementation, it would be between 0.6 and 1.8 with supplementation depending on the popultation.

Again this is just a prior and as we learn more about the virus we might discover that Covid is different from other respiratory virus and Vitamin D has no real effect. 
However, so far [almost everything](https://github.com/GShotwell/vitamin_d_covid) we've learned has supported a conenction between vitamin D and Covid outcomes:

- A re-analysis of [107 Swiss blood samples](https://www.mdpi.com/2072-6643/12/5/1359/htm) found that PCA positive patients had 25-hydroxyvitamin D concentrations half that of PCA negatives. This finding held after stratifying for age and gender.  
- A [restrospective study](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3571484) in the Philippines found a significant association between vitamin D status and severe Covid-19 infections.
- An [Indonesian](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3585561) study of 712 people found that after controlling for age, gender, and co-morbidity, vitamin D deficiency was associated with a twelve times higher risk of death.
- Hospitalized male Covid-19 patients [were found](https://www.medrxiv.org/content/10.1101/2020.05.01.20079376v1) to have lower vitamin D levels than controls
- A study out of [Northwestern University](https://www.medrxiv.org/content/10.1101/2020.04.08.20058578v3.full.pdf) found that the vitamin D status of a country's elderly population was associated with the number of severe cases of Covid-19 in that country

None of these data points are particular strong, but they all point in the same direction and reinforce the prior expectation that vitamin D reduces incidence of Covid infection in the same way it does other respiratory conditions. 

# Risk of action and inaction

What happens if we supplement everyone with vitamin D and it doesn't help Covid patients one bit?

Vitamin D supplementation is extremely safe, the above meta-analysis found that supplementation didn't increase death or adverse events from any cause. 
Even if widespread vitamin D supplementation doesn't prevent Covid, it will still prevent _other respiratory conditions_ which is positive.
In the scenario where supplementation is not helpful it would prevent infection and hospital visits from non-Covid pnemonia, as a result there would be more ventilators and hospital resources for the pandemic.
Similarly it probably improves Covid outcomes indirectly because people who get Covid-19 after an unrelated respiratory ailment probably do worse. 

So the upside of vitamin D supplementation is a significant reduction in Covid-19 transmission and mortality. 
The downside is that we don't change Covid-19 rates at all, but reduce the rate of other respiratory conditions, which in turns helps our Covid-19 response.   
The upside is large, the downside is negligible therefore we should supplement with Vitamin D. 

This argument should sound familliar because it's exactly the same one that convinced so many people to wear homemade masks.
There are no clinical trials testing mask use among the general population, 
and indeed it's pretty difficult to imagine such a trial ever being done. 
There aren't even any observational studies of populations in real world conditions, 
all we have are a few labratory studies that show that masks can trap viral particles, and an association between mask-wearing countries and low Covid rates. 
Despite that, we have a prior expectation that since Covid mostly comes out of people's mouths, 
covering those mouths probably has some protective effect and,
since mask use is basically risk-free, any protective effect is worth the effort.
The clinical data on vitamin D is quite a lot stronger than universal masking because we have high quality RCTs that show that vitamin D protects against other respiratory viruses, 
and we have at least a few observational studies that show low vitamin D patients are at a high risk for severe Covid.
Even if vitamin D has no effect on Covid, we can be fairly sure that supplementation will free up health care resources by reducing other illnesses, 
and we can be very sure that it won't cost much or cause any harm. 

# Recommendations

There are a number of clinical trials for Vitamin D in Covid-19, so we'll know more about this topic over the next six months. 
I think it is a mistake to wait for the results of these trials because there's substantial upside to supplementation, and no real downside. 
In particular :

- If you are vulnerable to low vitamin D you get safe sun exposure or should start supplementing with vitamin D
- Facilities such as prisons, long term care facilities, and hospitals where sun exposure is limited should provide vitamin D supplementation
- Governments develop vitamin D plans for winter months. Argentina and New Zealand should start doing this now, Canada and Europe should plan for the fall. 
- Health care workers, [especially people of colour](http://www.pulsetoday.co.uk/news/bapio-addresses-nhs-employers-over-high-mortality-rates-among-bame-staff/20040687.article), should be tested and treated for vitamin D deficiency

Again, these things may do nothing but reduce the number of non-Covid respiratory infections. 
But as interventions go, they are cheap, safe, and easy to implement. 
If there's any chance they can reduce Covid transmission, they are worth doing.  

